Cerebral oxygenation manuscript figures

### 2021-11-05 final_figures.Rmd
scripts, no edits

---
title: "Dysregulation of Cerebral Hemoglobin Content in Malawian Children with Malaria"
author: "Rachel Smith"
date: "`r Sys.Date()`"
output: html_document
---

@@

```{r setup, include = FALSE}

@@

library(tidyverse)
library(DescTools)
library(janitor)
library(FSA)
library(factoextra)
library(ggrepel)
library(ggpubr)
library(tidymodels)
library(vip)
library(scales)
library(reshape)
library(extrafont)
library(grDevices)

load("glm_model.Rdata")
load("results.Rdata")
load("roc_curves.Rdata")
load("dfa_example.Rdata")
load("dfa_results.Rdata")
load("all_dfa_results.Rdata")

@@

### Table 1: Summary of clinical data by patient group

@@

### Figure 1: Cerebral tissue hemoglobin concentration and oxygen saturation

@@

# extract variables of interest
voi <- df_results %>% colnames() %>% .[3:8]

# select from the long df
df_hbox_voi_long <- df_hbox_long %>% 
  dplyr::filter(variable %in% voi) %>% 
  mutate(variable = factor(variable, levels = c("avg_Hb_o2sat", "avg_Hb_conc",
                                                "Hb_o2sat_overall_alpha", "Hb_conc_overall_alpha",
                                                "Hb_o2sat_second_alpha", "Hb_conc_second_alpha")))
# create color vector
my_col <- c("CM" = "#00BFC4", "HC" = "#7CAE00", "UM" = "#F8766D")

# plot
df_hbox_voi_long %>% dplyr::filter(!(grepl("alpha", variable))) %>%

  ggplot(aes(x = Status, y = value)) +
  geom_violin(aes(color = Status)) +
  geom_jitter(position = position_jitter(0.2), shape = 1, size = 2) +
  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +
  facet_wrap(~variable, scales = "free_y") +

  scale_color_manual(values = my_col) +
  theme_bw() +
  theme(legend.position = "none")  +

  theme(text = element_text(family = "Arial"),
        strip.text = element_text(size = 15),
        axis.title = element_text(size = 15),
        axis.title.y = element_blank(),
        axis.text = element_text(size = 15))

@@

```{r fig1.pvals, eval = FALSE}

# hemoglobin concentration
kruskal.test(avg_Hb_conc ~ Status, data = df_hbox)
DescTools::DunnTest(avg_Hb_conc ~ Status, data = df_hbox, method = "BH")

# hemoglobin oxygen saturation
kruskal.test(avg_Hb_o2sat ~ Status, data = df_hbox)

@@

**Figure 1.** Shown are the variables measured by Oxiplex, plotted by patient status. Cerebral tissue hemoglobin concentration levels were different between cerebral malaria patients and the other two groups, but did not differ between uncomplicated malaria patients and healthy volunteers (Kruskal-Wallis adjusted p-value = 5.0x10^-7^; Dunn Test post-hoc adjusted p-values: CM vs HC = 3.2x10^-6^, CM vs UM = 2.8x10^-4^, HC vs UM = 0.87). Levels of cerebral hemoglobin O~2~ saturation did not differ among groups (Kruskal-Wallis adjusted p-value = 0.47).

### Figure 2: Detrended fluctuation analysis results

df_hbox_voi_long %>% dplyr::filter(grepl("second_alpha", variable)) %>%

  ggplot(aes(x = Status, y = value)) +
  geom_violin(aes(color = Status)) +
  geom_jitter(position = position_jitter(0.2), shape = 1, size = 2) +
  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +

  facet_wrap(~variable, scales = "free_y") +

  ylim(0, 1.5) +
  scale_color_manual(values = my_col) +
  theme_bw() +

  theme(text = element_text(family = "Arial"),
        strip.text = element_text(size = 15),
        axis.title = element_text(size = 15),
        axis.title.y = element_blank(),
        axis.text = element_text(size = 15),
        legend.position = "none")

@@

```{r fig2.pvals, eval = FALSE}

# hemoglobin concentration alpha
kruskal.test(Hb_conc_second_alpha ~ Status, data = df_hbox)
DescTools::DunnTest(Hb_conc_second_alpha ~ Status, data = df_hbox, method = "BH")

# hemoglobin oxygen saturation alpha
kruskal.test(Hb_o2sat_second_alpha ~ Status, data = df_hbox)
DescTools::DunnTest(Hb_o2sat_second_alpha ~ Status, data = df_hbox, method = "BH")

@@

**Figure 2.** Shown are the alpha values associated with the hemoglobin concentration and oxygen saturation signals measured by Oxiplex. Differences among cerebral tissue hemoglobin concentration alpha values were significant by Kruskal-Wallis (adjusted p-value = 7.16x10^-4^), and differences between uncomplicated malaria patients and the other two patient groups were significant by post-hoc Dunn Test (adjusted p-values: CM vs HV = 0.10, CM vs UM = 4.7x10^-4^, HV vs UM = 0.03). Cerebral hemoglobin O~2~ saturation alpha values differed between cerebral malaria patients and the other two patient groups (Kruskal-Wallis adjusted p-value = 4.2x10^-5^; Dunn Test post-hoc adjusted p-values: CM vs HV = 7.5x10^-4^, CM vs UM = 3.5x10^-4^, HV vs UM = 0.32).

### Figure 3: Logistic regression model on CM vs UM

#### A. PCA
```{r fig3a, fig.height = 8}

# include variables of interest, remove healthy children from analysis

df_hbox_voi <- df_hbox %>% 
  dplyr::select(c("subject_id", "Status", "Hematocrit", "Lactate", "avg_Hb_conc", "Hb_conc_second_alpha", "avg_Hb_o2sat", "Hb_o2sat_second_alpha"))

  df_hbox_impute <- df_hbox_voi %>% 
    group_by(Status) %>% 
    mutate_at(3:ncol(.), f_impute) %>% 
    ungroup()

# remove healthy volunteers for logistic regression analysis
df_hbox_log_reg <- df_hbox_impute %>% dplyr::filter(Status != "HC")

# convert tibble to df
df_hbox_voi_df <- df_hbox_log_reg %>% 
  dplyr::select(-c(subject_id, Status)) %>% 
  as.data.frame()

# add rownames
rownames(df_hbox_voi_df) <- df_hbox_log_reg %>% .$subject_id

# compute PCA
my_pca <- prcomp(df_hbox_voi_df, scale = TRUE)

# define groups
groups <- as.factor(df_hbox_log_reg$Status)

@@

  theme_bw() +
  theme(text = element_text(family = "Arial"),
        axis.text = element_text(size = 20),
        axis.title = element_text(size = 25),
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 20))

@@

#### B. Confusion matrix

```{r fig3b_prep, eval = FALSE}

library(themis)

# select [HbT], [HbT] alpha, and hematocrit
  df_hbox_model <- df_hbox_log_reg %>% 
    dplyr::select(c(Status, Hematocrit, avg_Hb_conc, Hb_conc_second_alpha)) %>% 
    mutate(`1/10_Hb_conc_alpha` = Hb_conc_second_alpha*0.1, avg_Hb_conc_nm = avg_Hb_conc/1000) %>% 
    dplyr::select(-c(Hb_conc_second_alpha, avg_Hb_conc)) %>% 
    mutate(Status = as.factor(ifelse(Status == "CM", 1, 0)))

  df_hbox_model <- df_hbox_log_reg %>% 
    dplyr::select(c(Status, Hematocrit, avg_Hb_conc, Hb_conc_second_alpha)) %>% 
    mutate(`1.10_Hb_conc_alpha` = Hb_conc_second_alpha*0.1) %>% 
    mutate(`1.10_Hb_conc_alpha_z` = (Hb_conc_second_alpha - mean(Hb_conc_second_alpha)) / sd(Hb_conc_second_alpha), 
           Hb_conc_z = (avg_Hb_conc - mean(avg_Hb_conc)) / sd(avg_Hb_conc),
           Hematocrit_z = (Hematocrit - mean(Hematocrit)) / sd(Hematocrit)) %>% 
    dplyr::select(-c(Hb_conc_second_alpha, `1.10_Hb_conc_alpha`, avg_Hb_conc, Hematocrit)) %>% 
    mutate(Status = as.factor(ifelse(Status == "CM", 1, 0)))

# resample
set.seed(234)
hbox_boot <- bootstraps(df_hbox_model, strata = Status)#, times = 1000)

# recipe
hbox_rec <- recipe(Status ~ ., data = df_hbox_model) %>% 
  step_BoxCox(all_predictors()) %>% 
  step_nzv(all_predictors()) %>% 
  step_smote(Status) # upsample to deal with unbalanced sample sizes

# create model on 1000 different resamples
glm_res <- hbox_wf %>% 
  add_model(glm_spec) %>%
  fit_resamples(
    resamples = hbox_boot,
    metrics = metric_set(roc_auc, accuracy, sensitivity, specificity),
    control = tune::control_resamples(save_pred = TRUE, verbose = TRUE)
  )

# finalize model
hbox_final <- hbox_wf %>% 
  add_model(glm_spec) %>% 
  finalize_workflow(glm_best) %>%
  fit(df_hbox_model) %>%
  pull_workflow_fit()

# save objects so model doesn't rerun every time we knit
save(glm_res, hbox_final, file = "glm_model.Rdata")

@@

```{r fig3b_plot, fig.height = 8}

final_res_conf_mat <- glm_res %>% 
  conf_mat_resampled() %>% 
  mutate(Prediction = ifelse(Prediction == 1, "CM", "UM")) %>% 
  mutate(Truth = ifelse(Truth == 1, "CM", "UM"))

total_true_UM <- final_res_conf_mat %>% dplyr::filter(Truth == "UM") %>% .$Freq %>% sum()
total_true_CM <- final_res_conf_mat %>% dplyr::filter(Truth == "CM") %>% .$Freq %>% sum()

final_res_conf_mat %>% 
  mutate(total_true = ifelse(Truth == "CM", Freq/total_true_CM, Freq/total_true_UM)) %>% 
  mutate(`Percent occurrence` = round(total_true*100, 2)) %>%

  # plot
  ggplot(aes(x = Truth, y = Prediction, fill = `Percent occurrence`, label = `Percent occurrence`)) +
  geom_tile() +
  geom_text(size = 6) +
  scale_fill_gradient2(low = "#4575b4", high = "#d73027") +

  theme_bw() +
  theme(text = element_text(family = "Arial"),
        axis.title = element_text(size = 25),
        axis.text = element_text(size = 20),
        legend.title = element_text(size = 20),
        legend.text = element_text(size = 15))

#### C. ROC curve of final model compared to other iterations

```{r fig3c_prep, eval = FALSE}

@@

```{r fig3c_plot, fig.height = 8}

@@

ggplot() +
  geom_path(data = df_roc %>% filter(model == "final"), 
            aes(x = 1 - specificity, y = sensitivity, color = model), 
            size = 2.5) +
  geom_path(data = df_roc %>% filter(model != "final"), 
            aes(x = 1 - specificity, y = sensitivity, color = model), 
            size = 1) +
  geom_abline(lty = 3) +
  labs(x = "False positive rate", y = "True positive rate") +
  theme_bw() +

  theme(text = element_text(family = "Arial"),
        axis.title = element_text(size = 25),
        axis.text = element_text(size = 20),
        legend.title = element_text(size = 20),
        legend.text = element_text(size = 15),
        legend.position = "top")

@@

**Figure 3.** Logistic regression model discriminating CM vs UM patients using cerebral tissue hemoglobin concentration, cerebral tissue hemoglobin concentration alpha, and hematocrit as the regressors **(A)** PCA of patients with CM and UM as determined by NIRS variables. Color indicates patient status, ellipses show median and 95% confidence interval on the cluster. **(B)** Confusion matrix representing the results of the logistic regression model averaged across 1000 bootstrap resamples. Percent occurrence indicates percentage times that the event (shown by the x- and y-axes) occurred.

### Table 2. Logistic regression model evaluation

#### A. Model metrics
```{r table2a}

table2a <- collect_metrics(glm_res) %>% 
  dplyr::select(.metric, mean, std_err) %>% 
  mutate(mean = mean*100, std_err = std_err*100) %>% 
  mutate(Metric = c("Accuracy", "ROC AUC", "Sensitivity", "Specificity")) %>% 
  dplyr::select(c("Metric", "mean", "std_err")) %>%

@@

table2a %>% knitr::kable(align = rep('c', 4))

@@

#### B. Coefficients
```{r table2b}

table2b <- hbox_final %>% 
  tidy(exponentiate = TRUE) %>% 
  mutate(estimate = ifelse(estimate < 10^-2, scientific(estimate, digits = 3), round(estimate, 2))) %>% 
  mutate(p.value = scientific(p.value, digits = 3),
         std.error = round(std.error, 2),
         statistic = round(statistic, 2)) %>% 
  dplyr::rename("Term" = "term", "Odds ratio" = "estimate", "Standard error" = "std.error", 
                "t-statistic" = "statistic", "p-value" = "p.value") %>% 
  as.data.frame()

table2b %>% knitr::kable(align = rep('c', 5))

```{r export_table2, eval = FALSE}

list(table2a, table2b) %>% write.csv("Outputs/table_2.csv")

**Table 2.** Evaluation of the best logistic regression model select from 1000.

### Figure 4: Relationships among hemoglobin concentration and hematocrit
```{r fig4, fig.height = 8}

df_hbox %>% filter(Status != "HC") %>%

  ggplot(aes(x = avg_Hb_conc, y = Hematocrit)) +
  geom_point(aes(color = Status), alpha = 0.7, size = 3) +

  xlab("Hemoglobin concentration (μM)") +
  ylab("Hematocrit (%)") +

  theme_bw() +
  theme(text = element_text(family = "Arial"),
        axis.text = element_text(size = 20),
        axis.title = element_text(size = 25),
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 20))

**Figure 4.**

### Figure 5: Cerebral malaria patient follow-up visits

```{r fig5_data}

# reorganize for plotting
df_results_all_plot <- df_brain_dfa_results %>% 
  dplyr::filter(!is.na(avg_Hb_conc_fu)) %>% 
  pivot_longer(cols = 3:ncol(.), names_to = "variable", values_to = "value") %>% 
  mutate(follow_up = ifelse(grepl("fu", variable), "follow-up", "initial visit")) %>% 
  mutate(follow_up = factor(follow_up, levels = c("initial visit", "follow-up")))

# descriptive statistics

  # averages

    # THC
    df_thc_compare <- df_results_all_plot %>% 
      dplyr::filter(variable %in% c("avg_Hb_conc", "avg_Hb_conc_fu"))

    thc_fu_pval <- wilcox.test(value ~ follow_up, data = df_thc_compare) %>% .$p.value

    # O2 sat
    df_o2sat_compare <- df_results_all_plot %>% 
      dplyr::filter(variable %in% c("avg_Hb_o2sat", "avg_Hb_o2sat_fu"))

    o2sat_fu_pval <- wilcox.test(value ~ follow_up, data = df_o2sat_compare) %>% .$p.value

  # alphas

    # THC
    df_thc_alpha_compare <- df_results_all_plot %>% 
      dplyr::filter(variable %in% c("Hb_conc_overall_alpha", "Hb_conc_fu_alpha"))

    thc_alpha_fu_pval <- wilcox.test(value ~ follow_up, data = df_thc_alpha_compare) %>% .$p.value

    # O2 sat
    df_o2sat_alpha_compare <- df_results_all_plot %>% 
      dplyr::filter(variable %in% c("Hb_o2sat_overall_alpha", "Hb_o2sat_fu_alpha"))

    o2sat_alpha_fu_pval <- wilcox.test(value ~ follow_up, data = df_o2sat_alpha_compare) %>% .$p.value

#### A: Cerebral hemoglobin oxygen saturation & hemolgobin concentration
```{r fig5a}

# create color vector
my_col_fu <- c("initial visit" = "#B80F0A", "follow up" = "black")

# plot
df_results_all_plot %>%

  dplyr::filter(grepl("avg_Hb", variable)) %>% 
  mutate(measurement = ifelse(grepl("conc", variable), "THC", "O2_sat")) %>% 
  mutate(measurement = factor(measurement, levels = c("O2_sat", "THC"))) %>% 
  mutate(pvals = ifelse(measurement == "THC", round(thc_fu_pval, 3), round(o2sat_fu_pval, 3))) %>% 
  mutate(ycoord = ifelse(measurement == "THC", 400, 100)) %>%

  ggplot(aes(x = follow_up, y = value)) +
  geom_violin(aes(color = follow_up)) +
  geom_point(shape = 1, size = 2) +
  stat_summary(fun = "median", geom = "crossbar", aes(x = follow_up, color = follow_up), size = 0.2, width = 0.5) +
  geom_path(aes(group = subject_id), lty = 2, alpha = 0.7) +

  # geom_segment(aes(x = 1, y = ycoord, xend = 2, yend = ycoord), size = 0.3) +
  # geom_segment(aes(x = 1, y = ycoord, xend = 1, yend = ycoord - .04*ycoord), size = 0.3) +
  # geom_segment(aes(x = 2, y = ycoord, xend = 2, yend = ycoord - .04*ycoord), size = 0.3) +
  # geom_text(x = 1.5, aes(y = ycoord + .025*ycoord, label = paste0("p = ", pvals))) +

  facet_wrap(~measurement, scales = "free_y") +
  labs(y = "", x = "") +
  scale_color_manual(values = my_col_fu) +

  theme_bw() +
  theme(legend.position = "none") +
  theme(text = element_text(family = "Arial"),
        strip.text = element_text(size = 15),
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 15),
        legend.position = "none")

```
No difference from initial visit to follow up - Hb_oxy pval: 0.90; Hb_tot pval: 0.32

#### B: Cerebral hemoglobin oxygen saturation & hemolgobin concentration alphas
```{r fig5b}

library(scales)

df_results_all_plot %>%

  dplyr::filter(!grepl("avg|second", variable)) %>% 
  mutate(measurement = ifelse(grepl("conc", variable), "THC", "O2_sat")) %>% 
  mutate(measurement = factor(measurement, levels = c("O2_sat", "THC"))) %>% 
  mutate(pvals = ifelse(measurement == "THC", round(thc_alpha_fu_pval, 3), round(o2sat_alpha_fu_pval, 3))) %>%

  ggplot(aes(x = follow_up, y = value)) +
  geom_violin(aes(color = follow_up)) +
  geom_point(shape = 1, size = 2) +
  stat_summary(fun = "median", geom = "crossbar", aes(x = follow_up, color = follow_up), size = 0.2, width = 0.5) +
  geom_path(aes(group = subject_id), lty = 2, alpha = 0.7) +

  geom_segment(aes(x = 1, y = 0.93, xend = 2, yend = 0.93), size = 0.3) +
  geom_segment(aes(x = 1, y = 0.93, xend = 1, yend = 0.89), size = 0.3) +
  geom_segment(aes(x = 2, y = 0.93, xend = 2, yend = 0.89), size = 0.3) +
  geom_text(x = 1.5, y = 0.96, aes(label = paste0("p = ", pvals))) +

  facet_wrap(~measurement) +

  ylim(0.45, 1.1) +
  labs(y = "alpha", x = "") +
  scale_color_manual(values = my_col_fu) +

  theme_bw() +
  theme(legend.position = "none") +
  theme(text = element_text(family = "Arial"),
        strip.text = element_text(size = 15),
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 15),
        legend.position = "none")

### Figure 6: Muscle hemoglobin concentration and oxygen saturation data from initial visit and follow-up

#### A. Muscle hemoglobin concentration and oxygen saturation

```{r fig6A}

# select from the long df
df_muscle_long <- df_muscle_dfa_results %>% 
  dplyr::select(c(Status, avg_Hb_conc_muscle, avg_Hb_o2sat_muscle)) %>% 
  pivot_longer(2:3, names_to = "variable") %>% 
  mutate(Status = factor(Status, levels = c("HC", "UM", "CM")),
         variable = factor(variable, levels = c("avg_Hb_o2sat_muscle", "avg_Hb_conc_muscle")))

# plot
df_muscle_long %>% dplyr::filter(Status != "?") %>%

  ggplot(aes(x = Status, y = value)) +
  geom_violin(aes(color = Status)) +
  geom_jitter(position = position_jitter(0.2), shape = 1, size = 2) +
  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +
  facet_wrap(~variable, scales = "free_y") +

  scale_color_manual(values = my_col) +
  theme_bw() +
  theme(legend.position = "none")  +

  theme(text = element_text(family = "Arial"),
        strip.text = element_text(size = 15),
        axis.title = element_text(size = 15),
        axis.title.y = element_blank(),
        axis.text = element_text(size = 15))

```{r fig6a.pvals, eval = FALSE}

# hemoglobin concentration
kruskal.test(avg_Hb_conc_muscle ~ Status, data = df_muscle_dfa_results)

# hemoglobin oxygen saturation
kruskal.test(avg_Hb_o2sat_muscle ~ Status, data = df_muscle_dfa_results)

```
No significant differences: Hb_oxy pval = 0.11 by KW; Hb_tot pval = 0.051 by KW

#### B. Muscle DFA results
```{r fig6B}

# select from the long df
df_muscle_dfa_long <- df_muscle_dfa_results %>% 
  dplyr::select(c(Status, muscle_Hb_conc_alpha, muscle_Hb_o2sat_alpha)) %>% 
  pivot_longer(2:3, names_to = "variable") %>% 
  mutate(Status = factor(Status, levels = c("HC", "UM", "CM")),
         variable = factor(variable, levels = c("muscle_Hb_o2sat_alpha", "muscle_Hb_conc_alpha")))

# plot
df_muscle_dfa_long %>% dplyr::filter(Status != "?") %>%

  ggplot(aes(x = Status, y = value)) +
  geom_violin(aes(color = Status)) +
  geom_jitter(position = position_jitter(0.2), shape = 1, size = 2) +
  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +

  facet_wrap(~variable) +
  ylim(0, 1.5) +

  scale_color_manual(values = my_col) +
  theme_bw() +
  theme(legend.position = "none")  +

  theme(text = element_text(family = "Arial"),
        strip.text = element_text(size = 15),
        axis.title = element_text(size = 15),
        axis.title.y = element_blank(),
        axis.text = element_text(size = 15))

```{r fig6b.pvals, eval = FALSE}

# hemoglobin concentration alpha
kruskal.test(muscle_Hb_conc_alpha ~ Status, data = df_muscle_dfa_results)

# hemoglobin oxygen saturation alpha
kruskal.test(muscle_Hb_o2sat_alpha ~ Status, data = df_muscle_dfa_results)

```
No significant differences: Hb_oxy pval = 0.08 by KW; Hb_tot pval = 0.15 by KW

#### C. CM follow-up muscle results
```{r fig6c_data}

# reorganize for plotting
df_muscle_fu_plot <- df_muscle_dfa_results %>% 
  dplyr::filter(!is.na(avg_Hb_o2sat_muscle_fu)) %>% 
  pivot_longer(cols = 3:ncol(.), names_to = "variable", values_to = "value") %>% 
  mutate(follow_up = ifelse(grepl("fu", variable), "follow-up", "initial visit")) %>% 
  mutate(follow_up = factor(follow_up, levels = c("initial visit", "follow-up")))

# descriptive statistics

  # averages

    # THC
    df_thc_muscle_compare <- df_muscle_fu_plot %>% 
      dplyr::filter(variable %in% c("avg_Hb_conc_muscle", "avg_Hb_conc_muscle_fu"))

    thc_muscle_fu_pval <- wilcox.test(value ~ follow_up, data = df_thc_muscle_compare) %>% .$p.value

    # O2 sat
    df_o2sat_muscle_compare <- df_muscle_fu_plot %>% 
      dplyr::filter(variable %in% c("avg_Hb_o2sat_muscle", "avg_Hb_o2sat_muscle_fu"))

    o2sat_muscle_fu_pval <- wilcox.test(value ~ follow_up, data = df_o2sat_muscle_compare) %>% .$p.value

  # alphas

    # THC
    df_thc_muscle_alpha_compare <- df_muscle_fu_plot %>% 
      dplyr::filter(variable %in% c("muscle_Hb_conc_alpha", "muscle_Hb_conc_fu_alpha"))

    thc_alpha_muscle_fu_pval <- wilcox.test(value ~ follow_up, data = df_thc_muscle_alpha_compare) %>% .$p.value

    # O2 sat
    df_o2sat_muscle_alpha_compare <- df_muscle_fu_plot %>% 
      dplyr::filter(variable %in% c("muscle_Hb_o2sat_alpha", "muscle_Hb_o2sat_fu_alpha"))

    o2sat_alpha_muscle_fu_pval <- wilcox.test(value ~ follow_up, data = df_o2sat_muscle_alpha_compare) %>% .$p.value

Muscle hemoglobin concentration and hemoglobin oxygen saturation
```{r fig6c_avgs}

df_muscle_fu_plot %>%

  dplyr::filter(grepl("avg_Hb", variable)) %>% 
  mutate(measurement = ifelse(grepl("conc", variable), "THC", "O2_sat")) %>% 
  mutate(measurement = factor(measurement, levels = c("O2_sat", "THC"))) %>% 
  mutate(pvals = ifelse(measurement == "THC", round(thc_muscle_fu_pval, 3), round(o2sat_muscle_fu_pval, 3))) %>% 
  mutate(ycoord = ifelse(measurement == "THC", 400, 100)) %>%

  ggplot(aes(x = follow_up, y = value)) +
  geom_violin(aes(color = follow_up)) +
  geom_point(shape = 1, size = 2) +
  stat_summary(fun = "median", geom = "crossbar", aes(x = follow_up, color = follow_up), size = 0.2, width = 0.5) +
  geom_path(aes(group = subject_id), lty = 2, alpha = 0.7) +

  # geom_segment(aes(x = 1, y = ycoord, xend = 2, yend = ycoord), size = 0.3) +
  # geom_segment(aes(x = 1, y = ycoord, xend = 1, yend = ycoord - .04*ycoord), size = 0.3) +
  # geom_segment(aes(x = 2, y = ycoord, xend = 2, yend = ycoord - .04*ycoord), size = 0.3) +
  # geom_text(x = 1.5, aes(y = ycoord + .025*ycoord, label = paste0("p = ", pvals))) +

  facet_wrap(~measurement, scales = "free_y") +
  labs(y = "", x = "") +
  scale_color_manual(values = my_col_fu) +

>>>>

---
title: "Cerebral oxygenation manuscript figures"
date: "`r Sys.Date()`"
output: html_document
---

@@

```{r setup, include=FALSE}

@@

# Load libraries
library(tidyverse)
library(DescTools)
library(janitor)
library(FSA)
library(factoextra)
library(ggrepel)
library(ggpubr)
library(tidymodels)
library(vip)
library(scales)
library(reshape)
library(extrafont)
library(grDevices)

# Load data
load("psd_plot_data.Rdata")
load("glm_model.Rdata")
load("results.Rdata")
load("roc_curves.Rdata")
load("dfa_example.Rdata")
load("dfa_results.Rdata")
load("all_dfa_results.Rdata")

@@

### Table 1. Baseline clinical characteristics

@@

### Fig 0. (optional) Power spectral density curves

```{r fig0}

# plot
f_plot_psd <- function(df, status, upper_lim) {

  df %>% 
    dplyr::filter(Status == status) %>% 
    dplyr::filter(freq > 0.01 & freq < upper_lim) %>% 
    ggplot(aes(x = freq, y = med_power)) +
    geom_line() +
    geom_vline(xintercept = 0.1, color = "red", lty = 2) +
    facet_grid(tissue ~ variable, scales = "free_y") +
    labs(x = "Frequency", y = "Median power across patients") +
    theme_bw()

## 0-1 Hz

  # HC
  f_plot_psd(df_psd, "HC", 1) + ggtitle("Healthy children")

  # UM
  f_plot_psd(df_psd, "UM", 1) + ggtitle("Uncomplicated malaria")

  # CM
  f_plot_psd(df_psd, "CM", 1) + ggtitle("Cerebral malaria")

## 0-3 Hz

  # HC
  f_plot_psd(df_psd, "HC", 3) + ggtitle("Healthy children")

  # UM
  f_plot_psd(df_psd, "UM", 3) + ggtitle("Uncomplicated malaria")

  # CM
  f_plot_psd(df_psd, "CM", 3) + ggtitle("Cerebral malaria")

### Fig 1. Hb_oxy and Hb_tot in the brain and muscle

@@

# extract variables of interest
voi <- df_all_dfa_results %>% colnames() %>% .[-(1:2)]

# select from the long df
df_hbox_voi_long <- df_hbox_long %>% 
  dplyr::filter(variable %in% voi & !grepl("fu", variable)) %>%

  mutate(tissue = as.factor(ifelse(grepl("muscle", variable), "muscle", "brain")),
         measurement = factor(ifelse(grepl("conc", variable), "Hb_tot", "Hb_oxy"), levels = c("Hb_oxy", "Hb_tot")),
         metric = case_when(

           grepl("overall_alpha|conc_alpha|o2sat_alpha", variable) ~ "alpha",
           grepl("second_alpha", variable) ~ "second_alpha",
           TRUE ~ "mean"

         )) %>% 
  dplyr::select(-variable)

# plot
df_hbox_voi_long %>% dplyr::filter(!(grepl("alpha", metric))) %>%

  ggplot(aes(x = Status, y = value)) +
  geom_violin(aes(color = Status)) +
  geom_jitter(position = position_jitter(0.2), shape = 1, size = 2) +
  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +
  facet_grid(measurement ~ tissue, scales = "free_y") +

  scale_color_manual(values = my_col) +
  labs(x = "") +
  theme_bw() +
  theme(legend.position = "none")  +

  theme(# text = element_text(family = "Arial"),
        strip.text = element_text(size = 15),
        axis.title = element_text(size = 15),
        axis.title.y = element_blank(),
        axis.text = element_text(size = 15))

@@

### Fig 2. Hb_oxy alpha and Hb_tot alpha in the brain and muscle

# plot
df_hbox_voi_long %>% dplyr::filter((grepl("second_alpha", metric))) %>%

  ggplot(aes(x = Status, y = value)) +
  geom_violin(aes(color = Status)) +
  geom_jitter(position = position_jitter(0.2), shape = 1, size = 2) +
  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +
  facet_grid(measurement ~ tissue) +

  ylim(0, 1.5) +
  scale_color_manual(values = my_col) +
  labs(x = "") +
  theme_bw() +
  theme(legend.position = "none")  +

  theme(# text = element_text(family = "Arial"),
        strip.text = element_text(size = 15),
        axis.title = element_text(size = 15),
        axis.title.y = element_blank(),
        axis.text = element_text(size = 15))

@@

### Fig 3. Logistic regression modeling of uncomplicated and cerebral malaria subjects

```{r fig3_data}

# include variables of interest

df_hbox_log_reg <- df_hbox %>% 
  dplyr::select(c("subject_id", "Status", "Hematocrit", "Lactate", "avg_Hb_conc", "Hb_conc_second_alpha", "avg_Hb_o2sat", "Hb_o2sat_second_alpha"))

  df_hbox_impute <- df_hbox_log_reg %>% 
    group_by(Status) %>% 
    mutate_at(3:ncol(.), f_impute) %>% 
    ungroup()

@@

#### A. Correlations of variables within subject groups

```{r fig3a, fig.show = "hold", out.width = "50%", fig.height = 12, fig.width = 15}

  m <- df %>%
    dplyr::filter(Status == status) %>% 
    dplyr::select(-c(subject_id, Status)) %>% 
    cor(method = "spearman")

  diag(m) <- 0

  m
}

# create run correlations of potential regressors
m_cor_um <- df_hbox_impute %>% f_calc_cor(status = "UM")

    ggplot(aes(x = X1, y = X2, fill = value, label = ifelse(value != 0, round(value, 2), NA))) + 
    geom_tile() +
    scale_fill_gradient2(low = "#4575b4", high = "#d73027") +
    geom_text(size = 9) +

    xlab("") +
    ylab("") +
    theme_bw() +
    theme(# text = element_text(family = "Arial"),
          axis.text.x = element_text(angle = 45, hjust = 1),
          axis.text = element_text(size = 35),
          legend.text = element_text(size = 40),
          legend.title = element_text(size = 50),
          legend.key.size = unit(2, 'cm'),
          plot.title = element_text(size = 30)
    )

  # extract legend
  p_leg <- m_cor_um %>% f_plot_cor() %>% get_legend()

  # create plots
  m_cor_um %>% f_plot_cor() + ggtitle("Uncomplicated malaria") + theme(legend.position = "none")

@@

#### B. Principal component analysis

```{r fig3b}
# remove healthy volunteers for logistic regression analysis
df_hbox_impute <- df_hbox_impute %>% dplyr::filter(Status != "HC")

# convert tibble to df
df_hbox_df <- df_hbox_impute %>% 
  dplyr::select(-c(subject_id, Status)) %>% 
  as.data.frame()

# add rownames
rownames(df_hbox_df) <- df_hbox_impute %>% .$subject_id

# compute PCA
my_pca <- prcomp(df_hbox_df, scale = TRUE)

# define groups
groups <- as.factor(df_hbox_impute$Status)

@@

  theme_bw() +
  theme(# text = element_text(family = "Arial"),
        axis.text = element_text(size = 20),
        axis.title = element_text(size = 25),
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 20))

@@

#### C. Coefficients
```{r table2c}

df_CI <- confint(hbox_final$fit) %>% exp() %>% as_tibble(rownames = "term")

table2c <- hbox_final %>%
  tidy(exponentiate = TRUE) %>%
  left_join(df_CI, by = "term") %>%
  mutate(estimate = ifelse(estimate < 10^-2, scientific(estimate, digits = 3), round(estimate, 2)),
         conf.low = ifelse(`2.5 %` < 10^-2, scientific(`2.5 %`, digits = 3), round(`2.5 %`, 2)),
         conf.high = ifelse(`97.5 %` < 10^-2, scientific(`97.5 %`, digits = 3), round(`97.5 %`, 2))) %>%
  mutate(p.value = scientific(p.value, digits = 3),
         statistic = round(statistic, 2)) %>%
  unite("95% CI", c(conf.low, conf.high), sep = ", ") %>%
  dplyr::select(c(term, estimate, `95% CI`, statistic, p.value)) %>% 
  dplyr::rename("Term" = "term", "Odds ratio" = "estimate",
                "t-statistic" = "statistic", "p-value" = "p.value") %>%
  as.data.frame()

table2c %>% knitr::kable(align = rep('c', 5))

@@

#### D. ROC curve of final model

```{r fig3d_prep, eval = FALSE}

@@

```{r fig3d_plot, fig.height = 8}

@@

ggplot() +
  geom_path(data = df_roc %>% filter(model == "final"), 
            aes(x = 1 - specificity, y = sensitivity, color = model), 
            size = 2.5) +
  geom_abline(lty = 3) +
  labs(x = "False positive rate", y = "True positive rate") +
  theme_bw() +

  theme(# text = element_text(family = "Arial"),
        axis.title = element_text(size = 25),
        axis.text = element_text(size = 20),
        legend.title = element_text(size = 20),
        legend.text = element_text(size = 15),
        legend.position = "none")

@@

```{r fig3d_inset}

fig3d_inset <- collect_metrics(glm_res) %>% 
  dplyr::select(.metric, mean, std_err) %>% 
  mutate(mean = mean*100, std_err = std_err*100) %>% 
  mutate(Metric = c("Accuracy", "ROC AUC", "Sensitivity", "Specificity")) %>% 
  dplyr::select(c("Metric", "mean", "std_err")) %>%

@@

fig3d_inset %>% knitr::kable(align = rep('c', 4))

@@

### Fig 4. Explanatory model of how to interpret hemoglobin concentration

See manuscript draft (figure created in PowerPoint)

## Supplement
### SFig 1. DFA analysis illustration
